**Title:** Increasing Chlorophyll *a* Amid Stable Nutrient Concentrations in Rhode Island Lakes and Reservoirs

### 2020-08-13 manuscript/ecosphere_resubmit/ecosphere_response.Rmd
making slow progress on revisions...

@@

>>>>

- **Response:** Another good suggestion.  We have addressed in our response to comment *1-5b* and *1-8*.  In particular, we have identified slope values that would indicate a meaningful ecological trend.  We use literature values as well as state water quality criteria (when available) to inform those values.  Additionally, we have added analysis to explore trends across trophic state.

@@

- **Response:** What we mean by this, is if sites with differing base lines (e.g. low nutrient site vs high nutrient site or cool water site vs warm water site) were included at different times, then site-specific anomalies are more appropriate for examining long-term trends becuase each of the sites will be re-scaled, relative to that sites long term mean.  For instance, comparing measured values from a cool water site that shows no trend which is measured mostly early in a sampling period to a warm water site without a trend measured late in a sampling period would likely show an increasing trend when in fact none existed.  Re-scaling these values relative to the mean sets both sites to a common baseline and would then not show a trend.  To further illustrate this, look at the two figures below.  The first figure shows simulated data from three sites.  All sites have no trend with site one being measured across the time period with values mostly in the middle of a range.  Site 2 has no trend with low values and is only in the first portion of the time period.  Site 3 has no trend with high values and is only in the later portion of the time period.  The second figure shows what the average yearly trend would look like if just looking at the values vs looking at the anomalies.  The measured values show a trend, when none should exist.  The anomalies correctly show no increasing trend.  We have added in these figures and text to the Discussion

```{r echo=TRUE}
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)
library(ggplot2)
library(hrbrthemes)
library(broom)

set.seed(0)
options(scipen=10000)

site1 <- tibble(site=rep("site 1",10),
                year = 2011:2020,
                rep_1 = rnorm(10,10,2.5),
                rep_2 = rep_1*rnorm(10, 1, 0.05),
                rep_3 = rep_1*rnorm(10, 1, 0.05)) %>%
  pivot_longer(rep_1:rep_3, names_to = "replicate", values_to = "values") %>%
  mutate(replicate = str_extract(replicate, "[0-9]"))

site2 <- tibble(site=rep("site 2",6),
                year = 2011:2016,
                rep_1 = rnorm(6,3,1),
                rep_2 = rep_1 * rnorm(6, 1, 0.05),
                rep_3 = rep_1 * rnorm(6, 1, 0.05)) %>%
  pivot_longer(rep_1:rep_3, names_to = "replicate", values_to = "values") %>%
  mutate(replicate = str_extract(replicate, "[0-9]"))

site3 <- tibble(site=rep("site 3",6),
                year = 2015:2020,
                rep_1 = rnorm(6, 17, 1),
                rep_2 = rep_1*rnorm(6, 1, 0.05),
                rep_3 = rep_1*rnorm(6, 1, 0.05)) %>%
  pivot_longer(rep_1:rep_3, names_to = "replicate", values_to = "values") %>%
  mutate(replicate = str_extract(replicate, "[0-9]"))

examp <- rbind(site1, site2, site3) %>%
  mutate(year = year(ymd(paste0(year,"01-01"))))

site_gg <- examp %>%
  ggplot(aes(x = year, y = values)) +
  geom_point() + 
  facet_grid(. ~ site) +
  theme_ipsum_rc()  +
  scale_x_continuous(labels = c(2012,2014,2016,2018,2020),
                       breaks = c(2012,2014,2016,2018,2020)) +
  labs(x = "Year", y = "Simulated Measurements")
site_gg

examp_site_summ <- examp %>% 
  group_by(site) %>%
  mutate(lt_mean = mean(values)) %>%
  ungroup() %>%
  group_by(site, year) %>%
  summarize(measured_value = mean(values),
            anomaly = mean(values-lt_mean)) %>%
  ungroup() %>%
  pivot_longer(measured_value:anomaly, "Method")

examp_yr_summ <- examp_site_summ %>%
  group_by(year, Method) %>%
  summarize(values = mean(value)) %>%
  ungroup()

meas_lm <- lm(values ~ year, 
              data = examp_yr_summ[examp_yr_summ$Method == "measured_value",]) %>% 
    tidy() %>%
    slice(2) %>%
    select(slope = estimate, p.value)
anom_lm <- lm(values ~ year, 
              data = examp_yr_summ[examp_yr_summ$Method == "anomaly",]) %>% 
    tidy() %>%
    slice(2) %>%
    select(slope = estimate, p.value)

simulated_trends_gg <- ggplot(examp_yr_summ, 
                              aes(x = year, y = values, color = Method)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("darkblue", "darkred"), 
                     labels = c("Anomaly", "Measured Values")) +
  labs(y = "Yearly Average Value", x = "Year", subtitle = paste0("Measured Values: slope = ", signif(meas_lm$slope, 2), 
                                      ", p-value = ", signif(meas_lm$p.value, 2), "\nAnomaly: slope = ", signif(anom_lm$slope, 2), 
                                      ", p-value = ", signif(anom_lm$p.value, 2))) +
  theme_ipsum_rc() +
  scale_x_continuous(labels = c(2012,2014,2016,2018,2020),
                       breaks = c(2012,2014,2016,2018,2020)) +
  theme(plot.subtitle = element_text(size=10, face="plain"))
simulated_trends_gg
```

### 2020-08-13 manuscript/ecosphere_resubmit/manuscript.Rmd
making slow progress on revisions...

We recognize that ecologically relevant change is system dependent and no single value will be universally appropriate.  We are using several values for this particular assessment, but other values could also be justified.  There are various ways to identify important temperature changes in lakes and reservoirs.  A policy based approach could examine agreements that attempt to limit temperature increases, such as the Paris Agreement, which aims to limit increases over the next century to 2°C, which would be 0.02°C per year.  Alternatively, we could look at more ecologically focused temperature changes.  For instance, Winder and Schindler [-@winder2004climate] saw disrupted trophic linkages between phytoplankton and zooplankton with 1.39°C over a 40 year data set, or 0.035°C per year.  For this study, we will use the 0.02°C per year as it would provide protections against other ecologically relevant changes (e.g. 0.035°C per year for trophic linkages) and would also be indicative of meeting an important policy goal. For changes in chlorophyll, total phosphorus, and total nitrogen an ecologically meaningful trend is one that would result in a change from an oligotrophic state to a mesotrophic state over the course of a century.  We use the values defined by the trophic state limits in Nürnberg [-@nurnberg1996trophic].  For chlorophyll the oligotrophic-mesotrophic limit is 3.5 μg/L and over a century that is 0.035 μg/L per year, the total phosphorus limit is 10 μg/L or 0.1 μg/L per year, and for total nitrogen the limit is 350 μg/L, or 3.5 μg/L.  As we have defined it, slope magnitudes in excess of these per year values will be considered to have an ecologically meaningful trend.

@@

The analysis approach we used here, site-specific anomalies, is not a novel method and does have a long history in the analysis of trends in climate [@hansen2006global; @hansen2010global; @jones1996calculating; @jones1999surface]. However, using it to examine water quality trends is a new application of the technique, as we could find little evidence of using it specifically for water quality trends.  We built on these methods and adapted them for use with long-term water quality trends.  While other methods are valid and robust [e.g., @Oliver_2017], we chose mean site-specific anomalies as they can provide readily interpretable results, especially for communicating to general audiences.  For instance, reporting the changes in anomalies allows us to look at changes in the original units.  With our analysis, the slope of the regression line for temperature suggests a mean yearly increase of `r signif(ww_temp_rl$slope, 2)` °C and the slope of the regression line for chlorophyll *a* shows a mean yearly increase of `r signif(ww_chla_rl$slope, 2)` µg/l.  Additionally, the site-specific anomalies are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (e.g., added later in a time period or removed).  Lastly, this analysis is only possible because of the availability of sound, long-term data on water quality in Rhode Island.  Without the URIWW data and the commitment and participation of more than 2500 volunteers over the years, our analyses would have been impossible.  Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in capturing and better understanding long term environmental trends.

@@

```{r year_num_table, echo=FALSE, results="asis"}
pander::pander(year_num_ww, style = "rmarkdown")
```
`r tab_num("year_n_ww")`

```{r year_num_table, echo=FALSE, results="asis"}
pander::pander(year_num_lagosne, style = "rmarkdown")
```
`r tab_num("year_n_lagosne")`

@@

>>>>

We recognize that ecologically relevant change is system dependent and no single value will be universally appropriate.  We are using several values for this particular assessment, but other values could also be justified.  There are various ways to identify important temperature changes in lakes and reservoirs.  A policy based approach could examine agreements that attempt to limit temperature increases, such as the Paris Agreement, which aims to limit increases over the next century to 2°C, which would be 0.02°C per year and similarly, the state of Rhode Island criteria for temperature increase is 2.2°C or 0.022°C per year over a century [@paris; @ri_wg_regs]. Alternatively, we could look at more ecologically focused temperature changes.  For instance, Winder and Schindler [-@winder2004climate] saw disrupted trophic linkages between phytoplankton and zooplankton with 1.39°C over a 40 year data set, or 0.035°C per year.  For this study, we will use the 0.02°C per year as it would provide protections against other ecologically relevant changes (e.g. 0.035°C per year for trophic linkages), meet local criteria, and would also be indicative of meeting an important policy goal. For changes in chlorophyll, total phosphorus, and total nitrogen an ecologically meaningful trend is one that would result in a change from an oligotrophic state to a mesotrophic state over the course of a century.  We use the values defined by the trophic state limits in Nürnberg [-@nurnberg1996trophic].  For chlorophyll the oligotrophic-mesotrophic limit is 3.5 μg/L and over a century that is 0.035 μg/L per year, for total nitrogen the limit is 350 μg/L, or 3.5 μg/L per year, and the total phosphorus limit is 10 μg/L or 0.1 μg/L per year.  This level for total phosphorus would also protect against As we have defined it, slope magnitudes in excess of these per year values will be considered to have an ecologically meaningful trend.

@@

The analysis approach we used here, site-specific anomalies, is not a novel method and does have a long history in the analysis of trends in climate [@hansen2006global; @hansen2010global; @jones1996calculating; @jones1999surface]. However, using it to examine water quality trends is a new application of the technique, as we could find little evidence of using it specifically for water quality trends.  We built on these methods and adapted them for use with long-term water quality trends.  While other methods are valid and robust [e.g., @Oliver_2017], we chose mean site-specific anomalies as they can provide readily interpretable results, especially for communicating to general audiences.  For instance, reporting the changes in anomalies allows us to look at changes in the original units.  With our analysis, the slope of the regression line for temperature suggests a mean yearly increase of `r signif(ww_temp_rl$slope, 2)` °C and the slope of the regression line for chlorophyll *a* shows a mean yearly increase of `r signif(ww_chla_rl$slope, 2)` µg/l.

The site-specific anomalies are also robust to variations in sampling effort and in the timing of inclusion of given sampling locations.  For instance, if a site included early in a time period had low values and site included late in the time period, then analyzing the measured values over this time period would show an increase whether or not one existed.  Using the site-specific anomalies rescales the values to a common framework and allows the real trend to be seen.  We illustrate this with simulated data that have no trend present (`r fig_num(name = "simulated_data", caption = "Simulated data showing example of sites with no trend present and a low site early in a time preiod, a high site late in a time period, and on site measured across the time period.", display = "cite")`).  The average of the yearly measured values shows a trend (slope = `r signif(meas_lm$slope, 2)` and p-value = `r signif(meas_lm$p.value, 2)`), whereas the site-specific anomalies correctly showed no trend (slope = `r signif(anom_lm$slope, 2)` and p-value = `r signif(anom_lm$p.value, 2)`, `r fig_num(name = "simulated_analysis", caption = "Analysis of simulated data with no trend present and a low site early in a time preiod, a high site late in a time period, and on site measured across the time period. Yearly average of the actual values is shown in red and yearly average of the site-specific anomalies is shown in blue.", display = "cite")`).

Lastly, this analysis is only possible because of the availability of sound, long-term data on water quality in Rhode Island.  Without the URIWW data and the commitment and participation of more than 2500 volunteers over the years, our analyses would have been impossible.  Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in capturing and better understanding long term environmental trends.

@@

```{r year_num_table_ww, echo=FALSE, results="asis"}
pander::pander(year_num_ww, style = "rmarkdown")
```
`r tab_num("year_n_ww")`

```{r year_num_table_lagos, echo=FALSE, results="asis"}
pander::pander(year_num_lagosne, style = "rmarkdown")
```
`r tab_num("year_n_lagosne")`

@@

![](../../figures/simulated_data.jpg)
`r fig_num("simulated_data")`

![](../../figures/simulated_analysis.jpg)
`r fig_num("simulated_analysis")`
